\textbf{Learning semi-Markovian DAGs with flow-based VAE}

### 2023-12-10 Tex/paper.tex
tex 수정

Here is the introduction of VAEs

@@

To compare the performance of our method with DAG-GNN, we conduct experiments on random graphs with 10, 20, and 30 nodes, assuming independent noise variables following Gaussian, Laplace, and exponential distributions. The results for the Gaussian case are shown in Figure \ref*{SHD_FDR_gaussian}. It can be observed that DAG-GNN outperforms our method in terms of SHD and FDR, considering the number of predicted edges. This is because DAG-GNN assumes that the noise variables follow a standard Gaussian distribution, making it more suitable for cases with independent Gaussian noise variables. In contrast, our method may be considered overparameterized in such cases.\\

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{fig/SHD_independence_normal.pdf}
        \caption{SHD with Gaussian noise}
        \label{SHD_gaussian}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{fig/FDR_independence_normal.pdf}
        \caption{FDR with Gaussian noise}
        \label{FDR_gaussian}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{fig/Predicted Edges_independence_normal.pdf}
        \caption{Number of predicted edges}
        \label{pred_edge_gaussian}
    \end{subfigure}
    \caption{SHD, FDR, and number of predicted edges with Gaussian noise}
    \label{SHD_FDR_gaussian}
\end{figure}

However, in cases where the noise variables follow Laplace or exponential distributions, our method consistently outperformed DAG-GNN in terms of both SHD and FDR. The results of the experiments are shown in Figure \ref*{SHD_FDR_laplace} and \ref*{SHD_FDR_exponential}. As the number of nodes increases, our method consistently demonstrates better performance compared to DAG-GNN in terms of both SHD and FDR.\\

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/SHD_independence_laplace.pdf}
        \caption{SHD with Laplace noise}
        \label{SHD_laplace}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/FDR_independence_laplace.pdf}
        \caption{FDR with Laplace noise}
        \label{FDR_laplace}
    \end{subfigure}
    \caption{SHD and FDR with Laplace noise}
    \label{SHD_FDR_laplace}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/SHD_independence_exponential.pdf}
        \caption{SHD with exponential noise}
        \label{SHD_exponential}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/FDR_independence_exponential.pdf}
        \caption{FDR with exponential noise}
        \label{FDR_exponential}
    \end{subfigure}
    \caption{SHD and FDR with exponential noise}
    \label{SHD_FDR_exponential}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{fig/comparison_indep_30_exponential_seed301.pdf}
    \caption{Example of estimated graphs with node size 30 and exponential noise}
    \label{example_graphs_indep}
\end{figure}

@@

To begin with, we set the proportion of bidirectional edges to 0.5 and compared the performance of our method with that of DAG-GNN. The results, as depicted in Figure \ref*{SHD_FDR_prop30}, demonstrate that our method outperforms DAG-GNN in terms of both SHD and FDR. Notably, with a node size of 20, our method exhibited stable training, resulting in a significant difference in SHD and FDR. Figure \ref*{example_graphs} presents the estimated graphs produced by each method. The graph estimated by DAG-GNN (third column) contains considerably more edges than the true graph (first column), whereas the graph estimated by our method (second column) has a number of edges much closer to that of the true graph.\\

@@

In this paper, we proposed a method to learn the structure of a semi-Markovian DAG using a flow-based VAE. We conducted experiments on simulated data and compared the performance of our method with that of DAG-GNN. The results demonstrate that our method outperforms DAG-GNN in terms of both SHD and FDR metrics, especially when the noise variables are dependent and the size of graph is large.

However, contrary to our initial expectations, the lower triangular matrix $\mathbf L$ learned in the IAF layer did not capture the covariance matrix of the actual noise variables. This seems to be due to the entanglement problem in the latent space, and resolving this issue remains a future research task. If the full covariance matrix of the noise variables could be found, we would be able to directly identify bidirectional edges in the graph, enabling more accurate learning of the semi-Markovian graph.

>>>>

VAE(Variational Auto-Encoder)\cite{kingma2013auto} is a neural network architecture that consists of a generative model and a variational inference. In VAE architecture, it is assumed that the continuous latent variable exists and its posterior probability is intractable. Therefore, variational inference is introduced to approximate the real posterior probability with ELBO(evidence lower bound).\\

Let $x$ be an observed variable and $z$ be a latent variable. $x$ is generated by some random process involving $z$. For parameter $\theta$, random variable $z$ is generated from prior distribution $p_{\theta}(z)$, then $x$ is generated from conditional distribution $p_{\theta}(x|z)$. And $p_{\theta}(x,z)$ becomes their parametric joint distribution. Given a dataset $\mathsf{X} = \{ x^{(1)}, ... , x^{(N)}\}$, maximum marginal likelihood learning is to maximize log likelihoods $\log{p_{\theta}(\mathsf{X})} = \sum_{i=1}^{N}{\log{p_{\theta}(x^{(i)})}}.$ However, marginal likelihood $p_{\theta}(x) = \int{p_{\theta}(z)p_{\theta}(x|z)dz}$ is intractable in general. Instead, we introduce the model parameter $\phi$ and a parametric inference model $q_{\phi}(z|x).$ Then we can reformulate the original log likelihoods with $q_{\phi}(z|x).$ \\
\begin{equation*}
	\begin{aligned}
		\log{p_{\theta}(x)} =  \mathbb{E}_{q_{\phi}(z|x)}\left[ \log{p_{\theta}(x)} \right] &= \mathbb{E}_{q_{\phi}(z|x)}\left[ \log{p_{\theta}(x)} + \log{p_{\theta}(z|x)} - \log{p_{\theta}(z|x)} \right] \\
		&= \mathbb{E}_{q_{\phi}(z|x)}\left[ \log{p_{\theta}(x, z)} - \log{q_{\phi}(z|x)} \right] + \mathbb{E}_{q_{\phi}(z|x)}\left[ \log{q_{\phi}(z|x)} - \log{p_{\theta}(z|x)} \right] \\
	\end{aligned}
\end{equation*}
\\

The second RHS term of the last equation is the KL divergence of the approximate posterior from the true posterior. Since KL divergence is non-negative, the first RHS term of the last equation can be expressed as $\mathcal{L}(\theta, \phi; x)$ and it is called variational lower bound.
\begin{equation*}
	\begin{aligned}
	\mathcal{L}(\theta, \phi; x) &= \mathbb{E}_{q_{\phi}(z|x)}\left[ \log{p_{\theta}(x, z)} - \log{q_{\phi}(z|x)} \right] \\
	\log{p_{\theta}(x)} &= \mathcal{L}(\theta, \phi; x) + D_{KL}( q_{\phi}(z|x) || p_{\theta}(z|x)) \geq \mathcal{L}(\theta, \phi ; x) \\
	\end{aligned}
\end{equation*}
\\
Therefore $\mathcal{L}(\theta, \phi; x)$ is the lower bound of the evidence, marginal likelihood. Under some condition of the distribution family $\{ q_{\phi} \}, $ maximizing $\mathcal{L}(\theta, \phi; x)$ with respect to $\{ \theta, \phi \}$ will simultaneously maximize log-likelihood $\log{p_{\theta}(x)}$ and minimize $ D_{KL}( q_{\phi}(z|x) || p_{\theta}(z|x))$. In VAE architecture, we set $\mathcal{L}(\theta, \phi; x)$ as objective function to be maximized.\\

The ELBO also can be written as 
\begin{equation*}
	\begin{aligned}
		\mathcal{L}(\theta, \phi ; x) &= -\mathbb{E}_{q_{\phi}(z|x)}\left[ \log{q_{\phi}(z|x)} - \log{p_{\theta}(z)} - \log{p_{\theta}(x|z)}  \right] \\
		&= -D_{KL}( q_{\phi}(z|x) || p_{\theta}(z)) + \mathbb{E}_{q_{\phi}(z|x)}\left[ \log{p_{\theta}(x|z)}  \right]. \\
	\end{aligned}
\end{equation*}

To estimate the lower bound, sampling method is used. However, Monte Carlo estimate of the expectation over $q_{\phi}(z|x)$ should be differentiable with respect to $\phi$ in gradient method. Reparametrization trick is introduced to solve this problem. It is often possible to express random variable $z \sim q_{\phi}(z|x) $ as a deterministic variable $z = g_{\phi}(\epsilon, x)$ for auxiliary variable $\epsilon \sim p(\epsilon)$. This reparametrization moves the uncertainty induced by $\phi$ into $\epsilon$ and makes the whole estimation differentiable.

@@

To compare the performance of our method with DAG-GNN, we conducted experiments on random graphs with independent noise variables. Specifically, we considered cases where the noise variables followed independent Gaussian, Laplace, and exponential distributions. The results of the experiments are shown in Figures \ref*{SHD_FDR_gaussian}, \ref*{SHD_FDR_laplace}, and \ref*{SHD_FDR_exponential}. In the case of Gaussian noise, despite using the original settings of DAG-GNN, the normalizing flow did not show superior performance compared to DAG-GNN. On average, for 10 nodes, our model and DAG-GNN achieved SHD values of 16.5 and 13, respectively. For 20 nodes, the SHD values were 24.7 and 22.1, and for 30 nodes, they were 21.8 and 18.4. However, in the case of non-Gaussian noise, our method consistently outperformed DAG-GNN in terms of both SHD and FDR as the number of nodes increased.

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{fig/SHD_independence_normal.pdf}
        \caption{SHD with Gaussian noise}
        \label{SHD_gaussian}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{fig/FDR_independence_normal.pdf}
        \caption{FDR with Gaussian noise}
        \label{FDR_gaussian}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=\textwidth]{fig/Predicted Edges_independence_normal.pdf}
        \caption{Number of predicted edges}
        \label{pred_edge_gaussian}
    \end{subfigure}
    \caption{SHD, FDR, and number of predicted edges with independent Gaussian noise}
    \label{SHD_FDR_gaussian}
\end{figure}

In cases where the noise variables follow Laplace or exponential distributions, our method consistently outperformed DAG-GNN in terms of both SHD and FDR. The results of the experiments are shown in Figure \ref*{SHD_FDR_laplace} and \ref*{SHD_FDR_exponential}. As the number of nodes increases, our method consistently demonstrates better performance compared to DAG-GNN in terms of SHD. The FDR values of our method are also lower than those of DAG-GNN, except for the case of 20 with exponential noise.\\

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/SHD_independence_laplace.pdf}
        \caption{SHD with Laplace noise}
        \label{SHD_laplace}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/FDR_independence_laplace.pdf}
        \caption{FDR with Laplace noise}
        \label{FDR_laplace}
    \end{subfigure}
    \caption{SHD and FDR with independent Laplace noise}
    \label{SHD_FDR_laplace}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/SHD_independence_exponential.pdf}
        \caption{SHD with exponential noise}
        \label{SHD_exponential}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.45\textwidth}
        \includegraphics[width=\textwidth]{fig/FDR_independence_exponential.pdf}
        \caption{FDR with exponential noise}
        \label{FDR_exponential}
    \end{subfigure}
    \caption{SHD and FDR with independent exponential noise}
    \label{SHD_FDR_exponential}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{fig/comparison_indep_30_laplace_seed201.pdf}
    \caption{Example of estimated graphs with node size 30 and exponential noise}
    \label{example_graphs_indep}
\end{figure}

@@

To begin with, we set the proportion of bidirectional edges to 0.3 and compared the performance of our method with that of DAG-GNN. The results, as depicted in Figure \ref*{SHD_FDR_prop30}, demonstrate that our method outperforms DAG-GNN in terms of both SHD and FDR. Notably, with a node size of 20, our method exhibited stable training, resulting in a significant difference in SHD and FDR. Figure \ref*{example_graphs} presents the estimated graphs produced by each method. The graph estimated by DAG-GNN (third column) contains considerably more edges than the true graph (first column), whereas the graph estimated by our method (second column) has a number of edges much closer to that of the true graph.\\

@@

In this paper, we proposed a method to learn the structure of a semi-Markovian DAG using a flow-based VAE. We conducted experiments on simulated data and compared the performance of our method with that of DAG-GNN. The results demonstrate that our method outperforms DAG-GNN in terms of both SHD and FDR metrics, especially when the noise variables have dependent structure and when the size of graph is large. In addition, our method produced a number of predicted edges closer to the true number of edges compared to DAG-GNN.\\

However, contrary to our initial expectations, the lower triangular matrix $\mathbf L$ learned in the IAF layer did not capture the covariance matrix of the actual noise variables. This seems to be due to the entanglement problem in the latent space, and resolving this issue remains a future research task. If the full covariance matrix of the noise variables could be found, it could be possible to directly identify bidirectional edges in the graph, enabling more accurate learning of the semi-Markovian graph.
